---
title: "Robustness check: dropping X1 and X3 from the LR / DT / RF models"
output: html_document
---

This script addresses the reviewer-style concern (raised independently by
HJMS and by a colleague reading the draft) that the models' predictive
performance might substantially reflect the fact that non-boundary points
in our dataset are generated as interval midpoints, which tend to have
much larger denominators than the boundary points themselves — i.e. that
$X_1$ (and to a lesser extent $X_3$) may be picking up a dataset
construction artefact rather than a genuine dynamical signature.

It refits the logistic regression, decision tree, and random forest models
under three predictor sets — the original 8 variables, with $X_1$ dropped,
and with both $X_1$ and $X_3$ dropped — using **the same train/validation/test
split and the same selected hyperparameters as in the paper**, and reports
how much predictive performance survives. It also computes a trivial
single-threshold baseline using only $\log_{10}(X_1)$, evaluated on the
same test split, as a sanity check on how much of the signal a naive
denominator-only rule already captures.

**Before running:** set your working directory to the folder containing
`mean_median_dataset.csv`, e.g.

```{r}
setwd("D:/022 - Bahan Ajar Genap + Pendek 2025-2026/Penelitian dan Pengabdian Masyarakat/Mandiri/Ko Jojo & Justin/Codings")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(caret)
library(rpart)
library(randomForest)
library(pROC)
library(dplyr)
```

## 1. Load data and reproduce the paper's train/validation/test split

```{r load-and-split}
set.seed(123)

raw <- read.csv("mean_median_dataset.csv", stringsAsFactors = FALSE)
names(raw) <- c("x1","x2","x3","x4","x5","x6","x7","x8",
                "x9","x10","x11","x12","y")

data8 <- raw[, c("x1","x3","x5","x6","x7","x9","x10","x11","y")]
data8$y <- as.factor(data8$y)

# Same stratified split as the main pipeline: stratify by y and by
# whether x5 != x9, 70/10/20 train/validation/test, seed 123.
data8$is_diff <- ifelse(data8$x5 != data8$x9, "Beda", "Sama")
data8$strata_group <- interaction(data8$y, data8$is_diff)

set.seed(123)
idx_train <- createDataPartition(data8$strata_group, p = 0.7, list = FALSE)
train_asli <- data8[idx_train, ]
temp       <- data8[-idx_train, ]

set.seed(123)
idx_val <- createDataPartition(temp$strata_group, p = 1/3, list = FALSE)
val_asli  <- temp[idx_val, ]
test_asli <- temp[-idx_val, ]

drop_helper_cols <- function(df) { df$is_diff <- NULL; df$strata_group <- NULL; df }
train_asli <- drop_helper_cols(train_asli)
val_asli   <- drop_helper_cols(val_asli)
test_asli  <- drop_helper_cols(test_asli)

cat("Train:", nrow(train_asli), "| Val:", nrow(val_asli), "| Test:", nrow(test_asli), "\n")
```

## 2. Helper functions

```{r helpers}
# Same log1p + scale preprocessing used for logistic regression only
proses_logreg <- function(df_input) {
  df <- df_input
  vars_skewed <- intersect(c("x1", "x3", "x7", "x11"), names(df))
  if (length(vars_skewed) > 0) df[vars_skewed] <- log1p(df[vars_skewed])
  vars_x <- setdiff(names(df), "y")
  df[vars_x] <- scale(df[vars_x])
  as.data.frame(df)
}

metrics_from_confusion <- function(pred, actual) {
  pred <- factor(pred, levels = c(0,1)); actual <- factor(actual, levels = c(0,1))
  tp <- sum(pred==1 & actual==1); tn <- sum(pred==0 & actual==0)
  fp <- sum(pred==1 & actual==0); fn <- sum(pred==0 & actual==1)
  acc <- (tp+tn)/(tp+tn+fp+fn)
  prec <- ifelse(tp+fp==0, NA, tp/(tp+fp))
  rec  <- ifelse(tp+fn==0, NA, tp/(tp+fn))
  f1   <- ifelse(is.na(prec)||is.na(rec)||(prec+rec==0), NA, 2*prec*rec/(prec+rec))
  c(Accuracy=acc, Precision=prec, Recall=rec, F1=f1)
}

# The three predictor sets to compare
variants <- list(
  "Original (8 vars)" = c("x1","x3","x5","x6","x7","x9","x10","x11"),
  "Drop X1"           = c("x3","x5","x6","x7","x9","x10","x11"),
  "Drop X1 and X3"    = c("x5","x6","x7","x9","x10","x11")
)
```

## 3. Refit all three models under all three variants

Decision tree uses the paper's selected $(N,K,A) = (50,25,4)$; random forest
uses the paper's selected $(P,f,K,A) = (100,4,50,5)$, i.e. `maxnodes = 2^5 = 32`.
Hyperparameters are held fixed across variants so the comparison isolates
the effect of removing the predictors, rather than re-optimising around them.

```{r refit-all, warning=FALSE}
results <- data.frame()
roc_list <- list()

for (vname in names(variants)) {
  vars <- variants[[vname]]
  cols <- c(vars, "y")

  ## Logistic regression
  tr <- proses_logreg(train_asli[, cols]); te <- proses_logreg(test_asli[, cols])
  lr_model <- glm(y ~ ., data = tr, family = binomial)
  lr_prob  <- predict(lr_model, newdata = te, type = "response")
  lr_pred  <- ifelse(lr_prob >= 0.5, 1, 0)
  lr_roc   <- roc(response = te$y, predictor = lr_prob, levels = c(0,1), quiet = TRUE)
  m <- metrics_from_confusion(lr_pred, te$y)
  results <- rbind(results, data.frame(Model="Logistic Regression", Variant=vname,
                                        t(as.data.frame(m)), AUC=as.numeric(auc(lr_roc))))
  roc_list[[paste("LR -", vname)]] <- lr_roc

  ## Decision tree
  tr_dt <- train_asli[, cols]; te_dt <- test_asli[, cols]
  dt_model <- rpart(y ~ ., data = tr_dt, method = "class",
                     control = rpart.control(minsplit = 50, minbucket = 25, maxdepth = 4))
  dt_prob <- predict(dt_model, newdata = te_dt, type = "prob")[,2]
  dt_pred <- predict(dt_model, newdata = te_dt, type = "class")
  dt_roc  <- roc(response = te_dt$y, predictor = dt_prob, levels = c(0,1), quiet = TRUE)
  m <- metrics_from_confusion(dt_pred, te_dt$y)
  results <- rbind(results, data.frame(Model="Decision Tree", Variant=vname,
                                        t(as.data.frame(m)), AUC=as.numeric(auc(dt_roc))))
  roc_list[[paste("DT -", vname)]] <- dt_roc

  ## Random forest
  tr_rf <- train_asli[, cols]; te_rf <- test_asli[, cols]
  rf_model <- randomForest(y ~ ., data = tr_rf, ntree = 100, mtry = min(4, length(vars)),
                            nodesize = 50, maxnodes = 32, importance = FALSE)
  rf_prob <- predict(rf_model, newdata = te_rf, type = "prob")[,2]
  rf_pred <- predict(rf_model, newdata = te_rf, type = "class")
  rf_roc  <- roc(response = te_rf$y, predictor = rf_prob, levels = c(0,1), quiet = TRUE)
  m <- metrics_from_confusion(rf_pred, te_rf$y)
  results <- rbind(results, data.frame(Model="Random Forest", Variant=vname,
                                        t(as.data.frame(m)), AUC=as.numeric(auc(rf_roc))))
  roc_list[[paste("RF -", vname)]] <- rf_roc
}

print(results, row.names = FALSE)
```

## 4. Trivial single-threshold baseline (denominator only)

```{r trivial-baseline}
log_denom_train <- log10(train_asli$x1 + 1e-10)
log_denom_test  <- log10(test_asli$x1 + 1e-10)
y_train <- as.numeric(as.character(train_asli$y))
y_test  <- as.numeric(as.character(test_asli$y))

cands <- seq(min(log_denom_train), max(log_denom_train), length.out = 2000)
accs  <- sapply(cands, function(t) mean((log_denom_train <= t) == y_train))
best_t <- cands[which.max(accs)]

baseline_pred <- as.integer(log_denom_test <= best_t)
m <- metrics_from_confusion(baseline_pred, y_test)
cat("Trivial baseline threshold (log10 scale):", round(best_t, 3), "\n")
print(m)

results <- rbind(results, data.frame(Model="Trivial baseline (denominator only)",
                                      Variant="log10(X1) threshold",
                                      Accuracy=m["Accuracy"], Precision=m["Precision"],
                                      Recall=m["Recall"], F1=m["F1"], AUC=NA))
```

## 5. Save results table (CSV)

```{r save-csv}
write.csv(results, "robustness_check_results.csv", row.names = FALSE)
cat("Saved robustness_check_results.csv\n")
```

## 6. ROC comparison plots (PDF, one per model)

```{r roc-plots}
plot_roc_group <- function(model_prefix, filename, title) {
  pdf(filename, width = 6, height = 6)
  keys <- names(roc_list)[startsWith(names(roc_list), model_prefix)]
  cols <- c("black", "firebrick", "steelblue")
  plot(roc_list[[keys[1]]], col = cols[1], lwd = 2, main = title)
  for (i in 2:length(keys)) {
    plot(roc_list[[keys[i]]], col = cols[i], lwd = 2, add = TRUE)
  }
  legend("bottomright", legend = sub(paste0(model_prefix, " - "), "", keys),
         col = cols[seq_along(keys)], lwd = 2, cex = 0.85)
  dev.off()
}

plot_roc_group("LR", "roc_logistic_regression.pdf", "Logistic Regression: ROC by variant")
plot_roc_group("DT", "roc_decision_tree.pdf", "Decision Tree: ROC by variant")
plot_roc_group("RF", "roc_random_forest.pdf", "Random Forest: ROC by variant")

cat("Saved roc_logistic_regression.pdf, roc_decision_tree.pdf, roc_random_forest.pdf\n")
```

## Summary

Compare the `Accuracy` and `AUC` columns across the three `Variant` rows
for each model in `robustness_check_results.csv`. If performance holds up
reasonably well under "Drop X1" and "Drop X1 and X3", that is direct
evidence against the models merely detecting the dataset-construction
artefact. The trivial baseline row shows how much a naive
denominator-only rule achieves on its own, for comparison.
